¥section{Preliminaries}¥label{sec:prelim}

We collect the definitions and conventions used throughout the paper.
Standard references for universal algebra and equational
logic include Burris and Sankappanavar‾¥cite{burris1981course},
McKenzie, McNulty, and Taylor‾¥cite{mckenzie1987algebras},
and Taylor‾¥cite{taylor1979equational}.

¥subsection{Magmas and equational laws}

¥begin{definition}¥label{def:magma}
A ¥emph{magma} is a pair $(M, *)$ where $M$ is a nonempty set and
$* ¥colon M ¥times M ¥to M$ is a binary operation.  No axioms are
imposed on‾$*$.
¥end{definition}

We work in the first-order language $¥mathcal{L} = ¥{*¥}$ with a
single binary function symbol.  ¥emph{Terms} are built from a countable
set of variables $¥{x, y, z, w, ¥ldots¥}$ and the operation‾$*$
according to the usual recursive definition: each variable is a term,
and if $s$ and‾$t$ are terms then $s * t$ is a term.

¥begin{definition}¥label{def:eq-law}
An ¥emph{equational law} is a formal identity $s = t$ where $s$
and‾$t$ are terms in‾$¥mathcal{L}$.  A magma $(M,*)$ ¥emph{satisfies}
$s = t$ if for every assignment $¥sigma$ of the variables to elements
of‾$M$, the identity $¥sigma(s) = ¥sigma(t)$ holds in $(M,*)$.
¥end{definition}

¥begin{definition}¥label{def:implication}
Let $E_1$ and $E_2$ be equational laws.  We say that $E_1$
¥emph{implies} $E_2$, written $E_1 ¥models E_2$, if every magma
satisfying $E_1$ also satisfies‾$E_2$.
¥end{definition}

By Birkhoff's theorem‾¥cite{birkhoff1935structure}, $E_1 ¥models E_2$
if and only if $E_2$ can be derived from $E_1$ using the rules of
equational logic (reflexivity, symmetry, transitivity, substitution,
and congruence).

¥subsection{The equation corpus}

Following Bolan et al.‾¥cite{bolan2025equational}, we consider the
finite set $¥Eq$ of all equational laws with at most four occurrences
of‾$*$.  This set has $|¥Eq| = 4694$ elements, partitioned into $1415$
equivalence classes under mutual implication.  Of the
$4694^2 = 22{,}033{,}636$ ordered pairs, $8{,}178{,}279$ (approximately
$37.12¥%$) are true implications.

¥subsection{Structural features of equational laws}¥label{subsec:features}

We define several syntactic features of an equational law $E¥colon s = t$
that are used in our prediction rules.

¥begin{definition}¥label{def:structural}
Let $E¥colon s = t$ be an equational law.
¥begin{enumerate}[label=(¥roman*)]
¥item The ¥emph{variable count} $¥var(E)$ is the number of distinct
  variables appearing in‾$E$.
¥item The ¥emph{size} $¥siz(E)$ is the total number of occurrences
  of‾$*$ in both $s$ and‾$t$.
¥item The ¥emph{signature} of‾$E$ is the pair $(a,b)$ where $a$
  (resp.‾$b$) is the number of occurrences of‾$*$ on the left-hand
  side (resp.¥ right-hand side).
¥item We say $E$ is ¥emph{bare} if one side of $E$ is a single
  variable.  Equivalently, $E$ has signature $(0, b)$ or $(a, 0)$
  for some $a$ or‾$b$.
¥item When $E$ is bare with a single variable $x$ on one side, we say
  $x$ is the ¥emph{anchor variable}.  We say $E$ is ¥emph{lhs-var}
  if $x$ appears on the left: $x = t$.
¥end{enumerate}
¥end{definition}

Without loss of generality, we may assume bare equations are written
in lhs-var form $x = t$ by applying symmetry.

¥begin{definition}¥label{def:projections}
Let $E¥colon x = t$ be a bare, lhs-var equational law.
¥begin{enumerate}[label=(¥roman*)]
¥item We say $E$ has the ¥emph{left-occurrence} property,
  $¥Lx(E) = ¥textsc{true}$, if the anchor variable‾$x$ appears
  as the left child of the root of‾$t$.
¥item We say $E$ has the ¥emph{right-occurrence} property,
  $¥Rx(E) = ¥textsc{true}$, if the anchor variable‾$x$ appears
  as the right child of the root of‾$t$.
¥item We say $E$ has the ¥emph{right-projection} property,
  $¥RP(E) = ¥textsc{true}$, if the right-hand side‾$t$ can be
  written as $u * x$ for some term‾$u$ not containing‾$x$.
¥end{enumerate}
¥end{definition}

¥begin{definition}¥label{def:topshape}
The ¥emph{top shape} of a term $t$ records the structure of its
two outermost levels.  If $t = s_1 * s_2$ and each $s_i$ is either
a variable (denoted $v$) or a product $s_i = u * w$ (denoted $m$),
then the top shape of‾$t$ is one of $v$-$v$, $v$-$m$, $m$-$v$,
or $m$-$m$.  If $t$ is a single variable, its top shape is‾$v$.
¥end{definition}

¥subsection{Projection algebras}

¥begin{definition}¥label{def:proj-alg}
A magma $(M, *)$ is a ¥emph{left-projection algebra} if $a * b = a$
for all $a, b ¥in M$.  It is a ¥emph{right-projection algebra} if
$a * b = b$ for all $a, b ¥in M$.
¥end{definition}

Left- and right-projection algebras are among the simplest magmas.
Every equational law satisfied by a projection algebra is trivially
implied by the projection identity, so these algebras play a
distinguished role in the implication graph.
